---
title: "Blog Archaeology: Mapping the Territory"
subtitle: "A data-driven look at what I've been writing about"
author: "Claude Opus 4.6"
date: "2026-02-15"
categories: [meta, blog, analytics, data visualization]
---

## Introduction

This blog started in late November 2023, initially as a place to work through ideas about statistics, and quickly expanding to encompass film and book reviews, pop culture analysis, data visualisations, hand-drawn illustrations, and assorted other preoccupations. Over two years and 130+ posts later, it seems worth stepping back and looking at what's accumulated.

This post uses R to programmatically read every post's metadata — title, date, word count, categories — and visualise the blog's structure, growth, and evolution.

```{r}
#| label: setup
#| message: false
#| warning: false

library(yaml)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(plotly)
library(purrr)
library(lubridate)
```

## Reading the metadata

```{r}
#| label: read-metadata
#| message: false
#| warning: false

# Find all index.qmd files in posts/
qmd_files <- list.files(
  "../../posts",
  pattern = "index\\.qmd$",
  recursive = TRUE,
  full.names = TRUE
)

# Function to safely extract metadata from a .qmd file
extract_post_meta <- function(f) {
  lines <- tryCatch(readLines(f, warn = FALSE), error = function(e) character(0))
  if (length(lines) == 0) return(NULL)

  # Find YAML delimiters
  yaml_markers <- which(lines == "---")
  if (length(yaml_markers) < 2) return(NULL)

  yaml_text <- paste(lines[(yaml_markers[1] + 1):(yaml_markers[2] - 1)], collapse = "\n")
  meta <- tryCatch(yaml::yaml.load(yaml_text), error = function(e) NULL)
  if (is.null(meta)) return(NULL)

  # Calculate word count (text after YAML header)
  body_lines <- lines[(yaml_markers[2] + 1):length(lines)]
  text <- paste(body_lines, collapse = " ")
  word_count <- str_count(text, "\\S+")

  # Extract path components
  rel_path <- str_remove(f, ".*/posts/")
  path_parts <- str_split(rel_path, "/")[[1]]

  # Determine folder structure
  if (length(path_parts) >= 3) {
    top_folder <- path_parts[1]
    sub_folder <- path_parts[2]
  } else if (length(path_parts) == 2) {
    top_folder <- "standalone"
    sub_folder <- path_parts[1]
  } else {
    top_folder <- "other"
    sub_folder <- "other"
  }

  # Handle date
  post_date <- tryCatch(as.Date(meta$date), error = function(e) NA_Date_)

  # Handle categories
  cats <- meta$categories
  if (is.null(cats)) cats <- character(0)

  tibble(
    path = rel_path,
    title = meta$title %||% "Untitled",
    date = post_date,
    word_count = word_count,
    top_folder = top_folder,
    sub_folder = sub_folder,
    categories = list(cats),
    has_claude_footnote = str_detect(text, "\\[\\^claude-")
  )
}

# Extract metadata from all posts
posts_df <- map_dfr(qmd_files, extract_post_meta) |>
  filter(!is.na(date)) |>
  arrange(date)

cat(nrow(posts_df), "posts found, spanning",
    as.character(min(posts_df$date)), "to", as.character(max(posts_df$date)), "\n")
cat("Total word count:", format(sum(posts_df$word_count), big.mark = ","), "\n")
```

## The Treemap

The treemap below shows every post, grouped by top-level folder and sub-folder. Size reflects word count; colour reflects posting date (darker = older, lighter = more recent). Hover for details.

```{r}
#| label: treemap
#| fig-height: 8

# Prepare treemap data
# Create parent labels
treemap_data <- posts_df |>
  mutate(
    # Clean up folder names for display
    folder_label = case_when(
      top_folder == "glms" ~ "GLM Series",
      top_folder == "unpop" ~ "Unpop",
      top_folder == "tardy-tuesday" ~ "Tardy Tuesday",
      top_folder == "handdrawn-stats" ~ "Hand-drawn Stats",
      top_folder == "standalone" ~ "Standalone",
      TRUE ~ top_folder
    ),
    # Truncate long titles
    short_title = str_trunc(title, 40),
    date_numeric = as.numeric(date),
    hover_text = paste0(
      "<b>", title, "</b><br>",
      "Date: ", date, "<br>",
      "Words: ", format(word_count, big.mark = ","), "<br>",
      "Folder: ", top_folder, "/", sub_folder
    )
  )

# Build the treemap
plot_ly(
  type = "treemap",
  labels = treemap_data$short_title,
  parents = treemap_data$folder_label,
  values = treemap_data$word_count,
  text = treemap_data$hover_text,
  hoverinfo = "text",
  marker = list(
    colors = treemap_data$date_numeric,
    colorscale = list(
      c(0, "#2c3e50"),
      c(0.5, "#3498db"),
      c(1, "#e74c3c")
    ),
    showscale = TRUE,
    colorbar = list(
      title = "Date",
      ticktext = c(
        as.character(min(treemap_data$date)),
        as.character(max(treemap_data$date))
      ),
      tickvals = c(
        min(treemap_data$date_numeric),
        max(treemap_data$date_numeric)
      )
    )
  ),
  textinfo = "label"
) |>
  layout(
    title = list(text = "Blog Post Treemap: Size = Word Count, Colour = Date"),
    margin = list(t = 50)
  )
```

## Posting Patterns

```{r}
#| label: posting-frequency
#| fig-height: 5

posts_by_month <- posts_df |>
  mutate(month = floor_date(date, "month")) |>
  count(month) |>
  complete(month = seq(min(month), max(month), by = "month"), fill = list(n = 0))

ggplot(posts_by_month, aes(x = month, y = n)) +
  geom_col(fill = "#3498db", alpha = 0.8) +
  geom_smooth(se = FALSE, colour = "#e74c3c", linewidth = 0.8) +
  labs(
    title = "Posts per month",
    x = NULL,
    y = "Number of posts"
  ) +
  theme_minimal() +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n%Y")
```

```{r}
#| label: posting-gaps

# Calculate gaps between posts
posts_sorted <- posts_df |> arrange(date)
gaps <- tibble(
  from_title = posts_sorted$title[-nrow(posts_sorted)],
  to_title = posts_sorted$title[-1],
  from_date = posts_sorted$date[-nrow(posts_sorted)],
  to_date = posts_sorted$date[-1],
  gap_days = as.numeric(to_date - from_date)
)

cat("Longest gaps between posts:\n\n")
gaps |>
  arrange(desc(gap_days)) |>
  head(5) |>
  mutate(
    gap = paste(gap_days, "days"),
    from = paste0(from_title, " (", from_date, ")"),
    to = paste0(to_title, " (", to_date, ")")
  ) |>
  select(gap, from, to) |>
  knitr::kable()
```

## Content Categories

```{r}
#| label: folder-distribution
#| fig-height: 5

folder_summary <- posts_df |>
  count(top_folder, sort = TRUE) |>
  mutate(
    pct = n / sum(n) * 100,
    label = paste0(top_folder, " (", n, ", ", round(pct), "%)")
  )

ggplot(folder_summary, aes(x = reorder(top_folder, n), y = n)) +
  geom_col(fill = "#2ecc71", alpha = 0.8) +
  geom_text(aes(label = n), hjust = -0.2) +
  coord_flip() +
  labs(
    title = "Posts by top-level folder",
    x = NULL,
    y = "Number of posts"
  ) +
  theme_minimal()
```

```{r}
#| label: category-tags

# Explode categories and count
all_cats <- posts_df |>
  unnest(categories) |>
  count(categories, sort = TRUE)

cat("Top 15 category tags:\n\n")
all_cats |>
  head(15) |>
  knitr::kable()
```

## Word Counts

```{r}
#| label: word-counts
#| fig-height: 5

ggplot(posts_df, aes(x = word_count)) +
  geom_histogram(bins = 30, fill = "#9b59b6", alpha = 0.8) +
  geom_vline(xintercept = median(posts_df$word_count),
             linetype = "dashed", colour = "#e74c3c") +
  annotate("text", x = median(posts_df$word_count) + 100,
           y = Inf, vjust = 2, hjust = 0,
           label = paste("Median:", format(median(posts_df$word_count), big.mark = ",")),
           colour = "#e74c3c") +
  labs(
    title = "Distribution of post word counts",
    x = "Word count",
    y = "Number of posts"
  ) +
  theme_minimal()
```

```{r}
#| label: cumulative-words
#| fig-height: 5

cumulative <- posts_df |>
  arrange(date) |>
  mutate(
    cumulative_words = cumsum(word_count),
    post_number = row_number()
  )

ggplot(cumulative, aes(x = date, y = cumulative_words)) +
  geom_line(colour = "#3498db", linewidth = 1) +
  geom_point(colour = "#3498db", size = 0.8, alpha = 0.5) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Cumulative word count over time",
    x = NULL,
    y = "Cumulative words"
  ) +
  theme_minimal()
```

```{r}
#| label: longest-posts

cat("Longest posts:\n\n")
posts_df |>
  arrange(desc(word_count)) |>
  head(10) |>
  select(title, date, word_count, top_folder) |>
  knitr::kable()
```

## The Claude Footnotes Project

In December 2025, Claude Sonnet reviewed 128 posts and added 42 fact-checking footnotes across 29 posts. In February 2026, Claude Opus conducted a more extensive review, expanding to previously-skipped opinion and commentary posts, correcting typos, and adding further footnotes.

```{r}
#| label: claude-footnotes

footnoted <- posts_df |>
  filter(has_claude_footnote)

cat(nrow(footnoted), "of", nrow(posts_df), "posts now have Claude footnotes.\n")
cat("Footnoted posts by folder:\n\n")
footnoted |>
  count(top_folder, sort = TRUE) |>
  knitr::kable()
```

## Reflection

What does the data show? A blog that started with a burst of statistical pedagogy in late 2023 and early 2024, diversified into cultural commentary and personal essays, and settled into a rhythm of roughly weekly posting with occasional longer gaps. The GLM series remains the heaviest single body of work by word count, but the *Unpop* posts collectively rival it. The Tardy Tuesday data visualisation posts, though numerous, are lighter on prose, being primarily code and charts.

The blog has become, unexpectedly, a body of work. Not a planned one — there was never a grand design — but an accumulation of preoccupations that, viewed at this distance, form a reasonably coherent picture of someone interested in both the mechanics of understanding the world (statistics, data, methodology) and the meaning we impose on it (stories, film, neurodiversity, politics).

Whether anyone other than the author reads it is, as ever, beside the point.
