---
title: "Interactive Sliders with Crosstalk and Plotly"
author: 
  - "Jon Minton"
date: "2023-11-30"
code-fold: true
categories: [R, plotly, crosstalk]
---

## Introduction

Below is an example of creating a plotly chart with an interactive slider using crosstalk.

By default, the plot shows the proportion of datazones in a local authority that are in the 15% most deprived datazones in Scotland. (Using the 2020 SIMD). 

The slider allows different thresholds than the 15% default to be selected. 

## Data Downloading and Preparation

*To see the code itself, just click on the word 'code' to open up the block'.*

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(plotly)
library(crosstalk)
library(here)
```

```{r}
#| warning: false
#| message: false
if(!file.exists(here("simd_data.xlsx"))){
  download.file(
    url = "https://www.gov.scot/binaries/content/documents/govscot/publications/statistics/2020/01/scottish-index-of-multiple-deprivation-2020-data-zone-look-up-file/documents/scottish-index-of-multiple-deprivation-data-zone-look-up/scottish-index-of-multiple-deprivation-data-zone-look-up/govscot%3Adocument/SIMD%2B2020v2%2B-%2Bdatazone%2Blookup.xlsx",
    destfile = here("simd_data.xlsx"),
    mode = "wb"
  )
}

dta <- openxlsx::readWorkbook(here("simd_data.xlsx"), sheet = "SIMD 2020v2 DZ lookup data")

```

The code for the figure itself is below. It's quite a convoluted process. There's almost certainly neater ways of doing this. The main thing to keep in mind is *all the figures exist; just only one is visible at a time*.


```{r}
#| warning: false
#| message: false
# So let's construct a new aval containing the different x-y tuples given the threshold selected

calc_prop_deprived <- function(q, dta){
    dta %>% 
      group_by(HBname) %>% 
      summarise(prop_deprived = mean(pct_rank < q)) %>% 
      ungroup()
}

df_rank <- 
  dta %>% 
    select(HBname, SIMD2020v2_Rank) %>% 
    mutate(pct_rank = SIMD2020v2_Rank / max(SIMD2020v2_Rank))


shared_df <- tibble(
  dep_quants = seq(0.05, 0.95, by = 0.05)
) %>% 
  mutate(derived_props = map(dep_quants, calc_prop_deprived, dta = df_rank)) %>% 
  unnest(derived_props) %>% 
  mutate(undep_quants = 1 - dep_quants) 


# Now to put it in the structure, and set active for `dep_quants = 0.15`


unique_dep_quants <- unique(shared_df$dep_quants)
n_steps <- length(unique_dep_quants)

dep_vals <- list()
for (step in 1:n_steps){
  tmp <- 
    shared_df %>% 
      filter(dep_quants == unique_dep_quants[step]) %>% 
      select(HBname, prop_deprived) %>% 
      mutate(HBname = reorder(HBname, prop_deprived))
  
  dep_vals[[step]] <- list(
    visible = FALSE,
    name = paste0('Quantile: ', unique_dep_quants[step]),
    x=tmp$prop_deprived,
    y=tmp$HBname
    
  ) 
}

# 15% is the third list object 

dep_vals[3][[1]]$visible = TRUE

# Now visualise 

# create steps and plot all traces
dep_steps <- list()
fig <- plot_ly() 
for (i in c(3, 1, 2, 4:n_steps)) { # Start with 3 as this is 15% and this should determine the default HB order 
 fig <- add_bars(fig,x=dep_vals[i][[1]]$x,  y=dep_vals[i][[1]]$y, visible = dep_vals[i][[1]]$visible, 
                 name = dep_vals[i][[1]]$name, orientation = 'h', hoverinfo = 'x+y', color = I("gray"),
                 showlegend = FALSE) %>% 
   layout(
      title = list(
        text = glue::glue("Proportion of datazones in Health Boards at least this deprived")
      ),
      xaxis = list(
        title = "Proportion this deprived in Health Board",
        range = list(0, 1)
      ),
      yaxis = list(
        title = "Health Board"
      )
   )

  step <- list(args = list('visible', rep(FALSE, length(dep_vals))),
               method = 'restyle')
  step$args[[2]][i] = TRUE  
  step$label = unique_dep_quants[i]
  dep_steps[[i]] = step 
}  
#names(dep_steps) <- unique_dep_quants

fig <- fig %>%
  layout(sliders = list(list(active = 2,
                             currentvalue = list(prefix = "Deprivation: "),
                             steps = dep_steps)))

fig

```

As you can see, there's still some work to do regarding formatting. But it works! 


## Static 

For comparison, here's the same data used to produce a static plot

```{r}
# Now to put it in the structure, and set active for `dep_quants = 0.15`

df_15pc <- shared_df |> 
  filter(between(dep_quants, 0.149, 0.151)) |> 
  select(-dep_quants, -undep_quants)

df_15pc |>
  mutate(pct_deprived = 100 * prop_deprived) |> 
  ggplot(aes(y= pct_deprived, x = fct_reorder(HBname, pct_deprived))) + 
  geom_bar(stat = "identity") +
  geom_text(
    aes(
      label = ifelse(df_15pc$prop_deprived > 0, sprintf("%.1f", pct_deprived), "")
    ), 
    color = "white",
    hjust = 1, 
    nudge_y = -0.5
  ) + 
  coord_flip() + 
  labs(
    x = "Health Board",
    y = "Percent of datazones in 15% most deprived proportion of Scotland",
    title = "Percent of datazones in Health Board in 15% most deprived areas of Scotland",
    subtitle = "SIMD 2020"
  ) + 
  geom_hline(yintercept = 0)
```

