---
title: "Tidy Tuesday 27 Feb 2024: Leap Years"
author: 
  - "Myriam Scansetti"
  - "Antony Clark"
  - "Jon Minton"
  - "Nicoloas Christofidis"
  - "Brendan Clarke"
  - "Kennedy Owuso-Afriyie"
  - "Emu the cat"
date: "2024-02-28"
code-fold: false
warning: false
message: false
categories: [R, tidy tuesday, leap years]
---

## Introduction

The latest TidyTuesday dataset was on births, deaths and other historical events that occurred in leap years, i.e. those years that include 29 February (such as 2024!). Further details are [here](https://github.com/rfordatascience/tidytuesday/tree/master/data/2024/2024-02-27). 

Myriam led the session, and Antony provided additional code for performing text field analysis after the session. 

Also, Emu the cat had the following contribution to make:

> 43e'/;£@@@@@@@@@@.[^1]

[^1]: I don't think even regex can help us with this one. 

## The session

We started by loading some packages

```{r}
# Option 1: tidytuesdayR package 
## install.packages("tidytuesdayR")
## install.packages("waldo")
## install.packages("tidytext")
## install.packages("textdata")
 
library(tidytuesdayR)
library(tidyverse)
library(waldo)
library(tidytext)
library(textdata)
 
```

We then had two ways of loading the data, in this case three datasets. As usual I'm switching to the url-based approach for the blog post

```{r}
# tuesdata <- tidytuesdayR::tt_load('2024-02-27')
# ## OR
# tuesdata <- tidytuesdayR::tt_load(2024, week = 9)
 
# events <- tuesdata$events
# births <- tuesdata$births
# deaths <- tuesdata$deaths
 
# Option 2: Read directly from GitHub
 
events <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-02-27/events.csv')
births <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-02-27/births.csv')
deaths <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-02-27/deaths.csv')
 
```

We noticed the births data include mention of at least one Pope. We wanted to explore more and less robust ways of finding popes in the `births` and `deaths` dataset

We could start by just looking for whether the word `Pope` is in the person field of `births`
```{r}
str_detect(births$person, "Pope")

```

We then used a little expression to make the query not case sensitive:
```{r}
deaths %>% filter(str_detect(person, "(?i)Pope"))
```

Another approach is to use `ignore_case` in the `regex()` function:
```{r}
deaths %>% filter(str_detect(person, regex("pope", ignore_case = TRUE)))
```

The only persons with `pope` in their name appear to be actual popes, not people who just happen to have the letters 'pope' in their surname.


Next we looked at number of events by year. We used two tidyverse approaches to producing this, one using `group_by` and `summarise`, the other using `count`. 

```{r}
number_events <- events %>% 
  group_by(year) %>% 
  summarise(n= n())
number_events
```

```{r}
number_events_2 <- events %>% 
  count(year)
number_events_2
```

We then tried different comparator functions to see if they all agreed the contents were identical, with some mixed and confusing results:
```{r}
waldo::compare(number_events, number_events_2)
```

`waldo` says they are the same.
```{r}
identical(number_events, number_events_2)
```

`identical` says they are not identical

```{r}
setequal(number_events, number_events_2)
```

But `setequal` doesn't find differences


```{r}
all.equal(number_events, number_events_2)
```

All equal reports a number of differences, related to the attributes (metadata) between the two objects being compared. 

Curiouser and Curiouser... 

Now let's plot the number of events over time 

```{r}
number_events %>% 
  ggplot(aes(x = year, y = n))+
  geom_col()
```

We wanted to know if there was anyone who was both recorded as being born and dying in a leap year:

```{r}
person_bd <- births %>% 
  inner_join(deaths, by = "person")

person_bd
```

One person (born in Scotland!)

We then looked text analysis, and in particular sentiment analysis of the content of the descriptio field:

```{r}
births %>% 
  unnest_tokens(word, description) %>% 
  anti_join(get_stopwords()) %>% 
  left_join(get_sentiments("afinn"))
```

```{r}
events %>% 
  unnest_tokens(word, event) %>% 
  anti_join(get_stopwords()) %>% 
  left_join(get_sentiments("afinn"))
```

Here's the words in the `afinn` object with the highest (most positive) sentiment
```{r}
get_sentiments("afinn") %>% 
  arrange(desc(value)) 
```

And here's an exploration of average sentiment by (leap)year based on the events description field:

```{r}
events |>
  unnest_tokens(word, event) |>
  anti_join(get_stopwords()) |>
  right_join(get_sentiments("afinn")) |>
  group_by(year) |>
  summarise(mean_sentiment = mean(value)) |>
  ggplot(aes(x = year, y = mean_sentiment)) +
  geom_point() +
  geom_smooth()

```

## Antony's script

Load libraries
```{r}
library(tidyverse)
library(lubridate)
library(countrycode)
```
some extra data sets re nationalities 

```{r}
demonym <- readr::read_csv("https://raw.githubusercontent.com/knowitall/chunkedextractor/master/src/main/resources/edu/knowitall/chunkedextractor/demonyms.csv",
                           col_names = c("demonym","geography"))

demonym$demonym <- tolower(demonym$demonym)
demonym$geography <- tolower(demonym$geography)
country <- tibble(country=countrycode::codelist$country.name.en)
```

Load data
```{r}
# tuesdata <- tidytuesdayR::tt_load('2024-02-27')

# list2env(tuesdata,.GlobalEnv)

glimpse(events)
glimpse(births)
glimpse(deaths)

```


Which cohort of leap day births is most represented in Wikipedia's data? 

Are any years surprisingly underrepresented compared to nearby years? 

What other patterns can you find in the data?

how many popes?
```{r}
births %>% 
  mutate(is_pope = grepl("pope",tolower(paste(person,description)))) %>% 
  count(is_pope)

```


count births by century ----

```{r}
getCenturyCorrected <- function(year) {
  if (year %% 100 == 0) {
    century <- year / 100
  } else {
    century <- ceiling(year / 100)
  }
  return(century)
}

getCenturyCorrected(1900)
getCenturyCorrected(1901)
```


```{r}
births %>% 
  mutate(century=sapply(year_birth,getCenturyCorrected)) %>% 
  count(century)
```


do count() and summarise(n=n()) give identical dataframes? Not always ----
```{r}
x <- births %>% count(year_birth)
y <- births %>% group_by(year_birth) %>% summarise(n=n())

identical(attributes(x), attributes(y))

names(x)==names(y)

identical(
  x,
  y
)
```


a rough stab (clearly flawed) at parsing nationality ----
```{r}
births_nationality <-
  bind_rows(
    births %>%
      tidytext::unnest_tokens(word, description) %>%
      anti_join(tidytext::get_stopwords(), "word") %>%
      left_join(
        demonym,
        by = c(word = "geography"),
        relationship = "many-to-many"
      ) %>%
      left_join(demonym, by = "demonym"),
    
    births %>%
      tidytext::unnest_tokens(word, description) %>%
      anti_join(tidytext::get_stopwords(), "word") %>%
      left_join(demonym, c(word = "demonym")) %>%
      left_join(demonym, "geography", relationship = "many-to-many")
  )

births_nationality %>% count(geography) %>% arrange(-n)
```

Now a pretty wordcloud
```{r}

events %>% 
  unnest_tokens(word, event) %>% 
  anti_join(get_stopwords(),"word") %>% 
  count(word) %>% 
  {wordcloud::wordcloud(words = .$word, 
            freq = .$n, min.freq = 1,
            max.words = 20, random.order = FALSE, rot.per = 0.35, 
            colors = RColorBrewer::brewer.pal(8, "Dark2"))}

```

A neutral word should have a sentiment score of 0, not NA. Let's make that change...

```{r}

afinn_sentiments <- get_sentiments('afinn')

events %>% 
  unnest_tokens(word, event) %>% 
  anti_join(get_stopwords(),"word") %>% 
  left_join(afinn_sentiments,"word") %>% 
  # filter(!is.na(value)) %>% 
  replace_na(list(value=0)) %>% 
  mutate(century = sapply(year,getCenturyCorrected)) %>% 
  group_by(century) %>% 
  summarise(mean_sentiment = mean(value)) %>% 
  ggplot(aes(x=century,y=mean_sentiment))+geom_line()
```














