---
title: "Tidy Tuesday: Christmas films"
author:
    - "Tom Fowler"
    - "Nick Christofides"
    - "Andrew Saul"
    - "Jon Minton"
date: "2023-12-21"
---

A shorter and even tardier Tidy Tuesday this week, given we gave ourselves only half an hour rather than the usual hour to look over the most recent dataset. 

The dataset was about Christmas films. 

Our first question: is *Die Hard a Christmas film?*

Not according to [the methods used to produce the dataset](https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-12-12). If a film doesn't have Christmas or equivalent in its title, it's not coming in! 



```{r setup}
#| include: FALSE
library(tidyverse)
library(tidytuesdayR)
```

- loading

```{r Load}
tt <- tidytuesdayR::tt_load('2023-12-12')

df1 <- tt[[1]]
df2 <- tt[[2]]

```

 - 
- count of films by year
```{r}
df1 %>%
  count(year, sort = TRUE)


```

- how many films by year -plot with log on y axis
```{r}
df1 %>%
  count(year) %>%
  ggplot(aes(x = year, y = n))+
  geom_point()+
  #stat_smooth()+
  scale_y_log10()


```
- how many films by year -plot with log on y axis
- filter by 1960 onwards

```{r}
df1 %>%
  filter(year >= 1960) %>%
  count(year) %>%
  ggplot(aes(x = year, y = n))+
  geom_point()+
  #stat_smooth()+
  scale_y_log10()

```
- how many films by year -plot with log on y axis
- filter by 1960 onwards

```{r}
df1 %>%
  filter(year >= 1960) %>%
  count(year) %>%
  ggplot(aes(x = year, y = n))+
  geom_point()+
  stat_smooth(method = "lm")+
  scale_y_log10()

```

# questions

- how are they published? [cinema / streaming?]
- is it on imdb?
- full inclusion of 2023?
- are more recent films rubbish?


```{r}
df1 %>%
  
  group_by(year) %>%
  summarise(avg_rating = mean(average_rating)) %>%
  ggplot(aes(x = year, y = avg_rating))+
  geom_point()
```
- number of films vs avg rating
- fewer films may drive extreme values


- number of films vs avg rating

```{r}
df1 %>%
  
  group_by(year) %>%
  summarise(
    avg_rating = mean(average_rating), 
    n_films = n() ) %>%
  ggplot(aes(x = n_films, y = avg_rating))+
  geom_point()
```