---
title: "Part Five: Traversing the Likelihood Landscape"
author: "Jon Minton"
date: "2024-01-02"
code-fold: true
warning: false
message: false
categories: [statistics, R]
bibliography: references.bib
---

## Recap

The [last part of the series](../../intro-to-glms/lms-are-glms-part-04/index.qmd) ended by showing how, except for linear regression, the kinds of answers users need from models aren't in tables of model coefficients, but in model predictions. However, we only showed predictions as point estimates, giving the false and unreasonable impression that the model predicts with perfect accuracy. 

This post, start of the the second section of the series, provides the theoretical and methodological background necessary to produce model predictions which are more *honest*, providing predictions with uncertainty intervals relating both to uncertainty about how the model has been fit, and fundamental variability in the stochastic processes that the statistical models aim to simulate. It's likely to be the most challenging part of the blog series, but worth sticking with. 

## Aim 

In [the first part of this series](../../intro-to-glms/lms-are-glms-part-01/index.qmd), I stated that statistical model fitting, within the generalised model framework presented in @KinTomWit00, involves adjusting candidate values for elements of $\beta = \{\beta_0, \beta_1, ..., \beta_K \}$ such that the difference between what the model predicts given some predictor values, $Y_i | X_i$, and what has been observed alongside the predictors, $y_i$, is minimised on average[^1] in some way.

[^1]: If $Y_i$ is what the model predicts given observations $X_i$, and $y_i$ is the outcome observed to have occurred alongside $X_i$, then we can call $\delta_i = h(y_i, Y_i)$ the difference, or error, between predicted and observed value. The function $h(.,.)$ is typically the squared difference between predicted and observed values, $(Y_i - y_i)^2$, but could also in principle be the absolute difference $|Y_i - y_i|$. Term-fitting algorithms usually compare not any individual $\delta_i$, but a sum of these error terms $\delta$. The aim of the algorithm is to find the set of $\beta$ terms that is least wrong for the whole dataset $D$, rather than any specific row in the dataset $D_i$.  

The aim of this post is to show how this process is typically implemented in GLMs, using likelihood theory. 

## Bayes' Rule and Likelihood 

Statisticians and more advanced users of statistical models often divide themselves into 'frequentists' and 'Bayesians'. To some extent the distinction is really between 'improper Bayesians' and 'proper Bayesians', however, as Bayes' Rule is at the root of both approaches. [Bayes' Rule](https://en.wikipedia.org/wiki/Bayes%27_theorem) is:

$$
P(A|B) = \frac{P(B|A)P(A)}{P(B)}
$$

Note in the above the left hand side of the equation is $P(A|B)$ and the right hand side of the equation *includes* $P(B|A)$. To write it out as awkward prose, therefore, Bayes' Rule is a way of expressing **that given this** in terms of **this given that**. 

As with much of algebra, $A$ and $B$ are just placeholders. We could instead use different symbols instead, such as: 

$$
P(\tilde{\theta} | y) = \frac{P(y | \tilde{\theta})P(\tilde{\theta})}{P(y)}
$$


Likelihood theory offers a way of thinking about how good a model is in terms of its relationship to the data. According to @King98 (p. 59), it can be expressed as: 

$$
L(\tilde{\theta}| y) = k(y) P(y | \tilde{\theta}) 
$$

Or 

$$
L(\tilde{\theta} | y) \propto P(y | \tilde{\theta})
$$

Where $\tilde{\theta}$ is a proposed parameter or parameter combination for the model, and $y$ is the observed outcome.[^2]

[^2]: As @King98 (p. 59) describes it, "$k(y)$ is an unknown function of the data. Whereas traditional probability is a measure of *absolute* uncertainty ... the constant $k(y)$ means that likelihood is only a *relative* measure of uncertainty"

The important thing to note is that both Bayes' Rule and Likelihood Theory are ways of expressing **this given that** as a function of **that given this**. Specifically, the model given the data, as a function of the data given the model. [^3]


[^3]: Frequentist approaches can thus be considered a kind of 'improper Bayesian' approach by considering $k(y)$ in the Likelihood formula as a stand-in for $\frac{P(\tilde{\theta})}{P(y)}$ in Bayes' Rule. Roughly speaking, it's because of the improperness of treating the two terms as equivalent, and the relativeness of $k(y)$, that mean frequentist probability statements can't be interpreted as Bayesian probability statements. But thinking of the two terms as equivalent can be helpful for spotting the similarity between the two formulae. 

## Likelihood for linear regression 

When, many years ago, I completed [the course](https://scholar.harvard.edu/msen/classes/government-2001-advanced-quantitative-research-methodology-professor-gary-king) from this modelling framework is most associated, a hazing ritual employed near the start of the course was to require participants to derive the likelihood of different model specifications. However, I don't feel like hazing myself right now, so instead we can use the derivation shown on [slide 8 of these slides](https://scholar.harvard.edu/files/msen/files/sen_gov2001_section5.pdf): 

$$
L(\beta, \sigma^2 | y) = \prod{L(y_i | \mu_i, \sigma^2)}
$$

Where $\mu = X \beta$, $i$ indicates an observation in the data (a row of $X$ when $X$ is in matrix form), and $\prod$ indicates the likelihoods from each observation should be multiplied with each other to derive the overall likelihood for all observed data.

In practice the log Likelihood, rather than the likelihood itself, is used, because this allows calculation of a sum of terms ($\sum$) rather than product of terms ($\prod$), and the latter tends to be computationally easier to calculate. 

As we are interested only in how likelihood varies as a function of those model parameters we wish to estimate, $\theta = \{\beta, \sigma^2\}$, some of the terms in the log likelihood expression can be omitted, leaving us with:

$$
\log{L(\beta, \sigma^2 | y)} \doteq \sum{-\frac{1}{2}[\log{\sigma^2} + \frac{(y_i - X_i\beta)^2}{\sigma^2}]}
$$

For all the complexity of the above expression, at heart it takes three inputs:

- $\theta = \{\beta, \sigma^2\}$ : The candidate *parameters* for the model. 
- $y$ : the observed **response** value from the dataset $D$
- $X$ : the observed **predictor** values from the dataset $D$

And returns one value, the log likelihood $\log{L(.)}$. 

To reiterate, we can't change the data, but we can keep changing the candidate parameters $\theta$. Each time we do so, $\log{L(.)}$ will change too. 

*The aim of model calibration, in the Likelihood framework, is to maximise the Likelihood.* The parameter set that maximises the likelihood is also the parameter set that maximises the log likelihood. 

To continue the example from the slides, we can write out a function for calculating the log likelihood of standard linear regression as follows:

```{r}
#| echo: true
llNormal <- function(pars, y, X){
    beta <- pars[1:ncol(X)]
    sigma2 <- exp(pars[ncol(X)+1])
    -1/2 * (sum(log(sigma2) + (y - (X%*%beta))^2 / sigma2))
}
```

In the above, `pars` is (almost but not quite) $\theta$, the parameters to estimate. For standard linear regression $\theta = \{\beta, \sigma^2\}$, where $\beta = \{\beta_0, \beta_1, ..., \beta_k\}$, i.e. a vector of beta parameters, one for each column (variable) in $X$, the predictor matrix of observations; this is why $beta$ is selected from the first K values in `pars` where K is the number of columns in $X$. 

The last value in `pars` is used to derive the proposed $\sigma^2$. If we call this last value `eta` ($\eta$), then we can say $\sigma^2 = e^{\eta}$. So, whereas $\theta$ is a vector that 'packs' $\beta$ and $\sigma^2$ into a single ordered series of values, `pars` packs `eta` in place of $\sigma^2$. This substitution of `eta` for $\sigma^2$ is done to make it easier for standard parameter fitting algorithms to work, as they tend to operate over the full real number range, rather than just over positive values. 

In order to illustrate how the log likelihood function `llNormal` works in practice, let's construct a simple toy dataset $D$, and decompose $D = \{y, X\}$, the two types of data input that go into the `llNormal` function. 

```{r}
# set a seed so runs are identical
set.seed(7)
# create a main predictor variable vector: -3 to 5 in increments of 1
x <- (-3):5
# Record the number of observations in x
N <- length(x)
# Create a response variable with variability
y <- 2.5 + 1.4 * x  + rnorm(N, mean = 0, sd = 0.5)

# bind x into a two column matrix whose first column is a vector of 1s (for the intercept)

X <- cbind(rep(1, N), x)
# Clean up names
colnames(X) <- NULL

```

In the code above we have created $y$, a vector of nine observed responses; and $X$, a matrix of predictors with two columns (the number of variables for which $beta$ terms need to be estimated) and nine rows (the number of observations). 

Graphically, the relationship between x and y looks as follows:

```{r}
library(tidyverse)
tibble(x=x, y=y) |>
    ggplot(aes(x, y)) + 
    geom_point()

```

In this toy example, but almost never in reality, we *know* the correct parameters for the model. These are ${\beta_0 = 2.5, \beta_1 = 1.4}$ and $\sigma^2 = 0.25$. [^4] Soon, we will see how effectively we can use optimisation algorithms to recover these true model parameters. But first, let's see how the log likelihood varies as a function jointly of different candidate values of $\beta_0$ (the intercept) and $\beta_1$ (the slope parameter), if we already set $\sigma^2$ to `0.25`. 


[^4]: i.e. the square of the `sd` passed to `rnorm()` of 0.5


```{r}

candidate_param_values <- expand_grid(
    beta_0 = seq(-5, 5, by = 0.1),
    beta_1 = seq(-5, 5, by = 0.1)
)

feed_to_ll <- function(b0, b1){
    pars <- c(b0, b1, log(0.25))
    llNormal(pars, y, X)
}

candidate_param_values <- candidate_param_values |>
    mutate(
        ll = map2_dbl(beta_0, beta_1, feed_to_ll)
    )
```

```{r}
candidate_param_values |>
    ggplot(aes(beta_0, beta_1, z = ll)) + 
    geom_contour_filled() + 
    geom_vline(xintercept = 0) +
    geom_hline(yintercept = 0) +
    labs(
        title = "Log likelihood as a function of possible values of beta_0 and beta_1",
        x = "beta0 (the intercept)",
        y = "beta1 (the slope)"
    )
```

Looking at this joint surface of values, we can see a 'hotspot' where $\beta_0$ is around 2.5, and $\beta_1$ is around 1.4, just as we should expect. We can check this further by filtering candidate_param_values on the highest observed values of ll.

```{r}
candidate_param_values |> 
    filter(ll == max(ll))

```


## Summary 

We have now introduced the concepts of Bayes Rule, Likelihood, and log likelihood, then derived the log likelihood for standard linear regression. We then built a toy dataset where we know the true parameters, and looked at how the log likelihood varies as different $\beta$ parameters are proposed. We identified a 'hot spot' when the $\beta$ parameters proposed are close to the 'true values'. 

## Coming up

The [next part of this series](../lms-are-glms-part-06/index.qmd) shows how log likelihood functions tend to be used in practice, in conjunction with optimisation algorithms that (usally) arrive at good estimates of our coefficients in far fewer steps than we've used above.[^4]

[^4]: **Note from Claude:** Likelihood theory and maximum likelihood estimation (MLE) are foundational to both classical statistics and modern machine learning. In ML, what statisticians call "negative log-likelihood" is exactly what ML practitioners call "loss functions"—cross-entropy loss for classification and mean squared error for regression are both derived from likelihood theory. Stanford's CS229 (Machine Learning) by Andrew Ng covers this connection explicitly. For Python implementations, scipy.optimize.minimize() provides the same optimization functionality as R's optim(), while TensorFlow and PyTorch implement automatic differentiation to calculate gradients for these loss functions. @murphy2022probabilistic provides comprehensive coverage of how likelihood theory extends to modern ML methods. 
