---
title: "Post with code"
author: "Jon Minton"
date: "2023-11-28"
categories: [R]
---

This short post is intended to confirm that I can run and render R code within a Quarto blog post.

### Very simple example

Let's start off with some very simple base-R

```{r}
1 + 1
```

And of course let's not forget the obligatory

```{r}
statement <- "Hello World"

statement
```

### Graphs

Let's now look at a base-R graphic, again using a cliched example

```{r}
plot(mtcars$mpg ~ mtcars$wt)
```

### Some extensions

Let's now continue to be cliched, and load and use the tidyverse

```{r}
#| message: false
#| echo: false
#| warning: false
library(tidyverse)
```

```{r}
glimpse(mtcars)
```

```{r}
mtcars |> 
  group_by(carb) |> 
  summarise(
    mean_mpg = mean(mpg)
  ) |> 
  ungroup()
```

And to visualise

```{r}
mtcars |> 
  mutate(cyl = factor(cyl)) |> 
  ggplot(aes(x = wt, y = mpg, colour = cyl, group= cyl)) + 
  geom_point(aes(shape = cyl)) + 
  stat_smooth(se = FALSE, method = "lm") + 
  labs(
    x = "Weight", 
    y = "Miles per gallon"
  )
```

### Conclusion

So far, so good...
