---
title: "Tardy Tuesday: Olympics"
date: "2024-08-07"
author: 
    - "Andrew"
    - "Brendan"
    - "Abram"
    - "Kennedy"
    - "Imran"
    - "Gatz"
    - "Jon"
code-fold: true
warning: false
categories: [tidy tuesday, R, olympics]
---

The most recent Tidy Tuesday contained [various types of data on previous Olympics](https://github.com/rfordatascience/tidytuesday/tree/master/data/2024/2024-08-06). Andrew led the session, with some additional contributions from Brendan. 

## Andrew's script

Load packages and data
```{r}
library(tidyverse)

tt_list <- tidytuesdayR::tt_load("2024-08-06")

list2env(tt_list, envir = .GlobalEnv)
```

Take a glimpse:

```{r}
glimpse(olympics)
```

How many sports of which kinds?

```{r}
olympics %>% 
  count(sport, sort=T) 
```

Events within the sport of athletics

```{r}
olympics %>% 
  filter(sport == "Athletics") %>% 
  count(event, sort=T) 
```

A visualisation: we wanted to show how long different sports have been running:

```{r}
plot_obj <- 
  olympics %>% 
  filter(sport == "Athletics") %>% 
  group_by(event) %>% 
  summarise(Earliest_year = min(year),
            Latest_year = max(year),
            n = n()) %>%  
  arrange(desc(n)) %>% 
  mutate(year_range = Latest_year - Earliest_year+1,
         event = fct_reorder(event, year_range)) %>% 
  mutate(Sex = case_when(str_detect(str_to_lower(event), "women") ~ "Female",
                         str_detect(str_to_lower(event), "men") ~ "Male",
                         .default = "Other")) %>% 
  mutate(event_status = if_else(Latest_year == 2016, "Live", "Dead"))

plot_obj %>%
  slice(1:20) %>% 
  ggplot(aes(y=event, color = Sex))+
  geom_errorbarh(aes(xmin=Earliest_year, xmax=Latest_year, height=0))
```

What about just Female events?


```{r}
plot_obj %>%
#  slice(1:20) %>% 
  filter(Sex == "Female") %>% 
  ggplot(aes(y=event))+
  geom_errorbarh(aes(xmin=Earliest_year, xmax=Latest_year, height=0))+
  scale_x_continuous(breaks = seq(1920, 2020, by=10))
```

Slight modification to indicate whether events are still running:

```{r}
plot_obj %>%
  #  slice(1:20) %>% 
  filter(Sex == "Female") %>% 
  ggplot(aes(y=event))+
  geom_errorbarh(aes(xmin=Earliest_year, xmax=Latest_year, height=0,
                     linetype = event_status))+
  scale_x_continuous(breaks = seq(1920, 2020, by=10)) + 
    scale_linetype_manual(values = c("dashed", "solid"))

```


## Brendan's script

```{r}
olympics_dec <- olympics |>
    mutate(decade = round(year, -1)) |>
    mutate(bmi = weight / (height/100)^2) |>
    filter(!is.na(bmi)) |> 
    group_by(event, sex, decade) |>
    summarise(mean_bmi = mean(bmi), 
              min_bmi = min(bmi),
              max_bmi = max(bmi)) |>
    arrange(desc(mean_bmi))

olympics_dec |>
    distinct(event)

```

Have shotputters got heavier or lighter over time?

```{r}
olympics_dec |>
    filter(str_detect(event, "Shot")) |>
    add_count(event) |>
    filter(n > 3) |>
    ggplot(aes(y = mean_bmi, x = decade, color = event)) +
    geom_smooth() +
    geom_linerange(aes(ymin = min_bmi, ymax = max_bmi)) +
    facet_wrap(~sex, ncol = 1) +
    theme(legend.position = "none")
```

Maybe a little heavier, but for men it seems to have plateaued. For women it looks like a trend towards increased heaviness is still continuing.
