---
title: "Tidy Tuesday 23 April 2024: Space objects"
author: 
    - "Myriam Scansetti"
    - "Brendan Clarke"
    - "Tricia Govindasamy"
    - "Jon Minton"
    - "Andrew Saul"
    - "Gatz Osorio"
    - "Kennedy Owusu-Afriyie"
date: "2024-04-26"
code-fold: true
warning: false
message: false
categories: [tidy tuesday, space]
---

The most recent [TidyTuesday dataset](https://github.com/rfordatascience/tidytuesday/tree/master/data/2024/2024-04-23) was taken from [Our World In Data](https://ourworldindata.org/grapher/yearly-number-of-objects-launched-into-outer-space), and showed the number of objects launched into space by different entities by year. Myriam led this week's session. 

## Data preparation and exploration

```{r}
# Option 1: tidytuesdayR package 
#install.packages("tidytuesdayR")
library(tidyverse)
library(janitor)
```

```{r, cache = TRUE}
tuesdata <- tidytuesdayR::tt_load('2024-04-23')
 
outer_space_object <- tuesdata$outer_space_objects
unique(outer_space_object$Entity)
```

We wanted to know what the entities referred to:
```{r}
outer_space_object %>% 
  filter(!is.na(Code)) %>% 
  pull(Entity) %>% 
  unique()
```

Generally, when an entity has a 'code', it is a country (though contains some other aggregations, like whole world, as well)
```{r}
outer_space_object %>% 
  filter(!is.na(Code)) %>% 
  distinct(Entity) %>% 
  arrange(desc(Entity))
```

We decided to use `janitor::clean_names` to avoid having to remember the case of columns. 
```{r}
outer_space_objects <- outer_space_object %>% 
  clean_names()
outer_space_objects
```

## Visualisation 

How many objects released into space by the world by year? 

```{r}
outer_space_objects %>% 
  filter(entity == "World") %>% 
  group_by(year) %>% 
  summarise(num_objects = sum(num_objects)) %>% 
  ungroup() %>% 
  ggplot(aes(x = year, y =num_objects))+
  geom_line()
```

We thought maybe there's been, and been periods of, exponential growth, so looked at this with a log y scale too:

```{r}
outer_space_objects %>% 
  filter(entity == "World") %>% 
  group_by(year) %>% 
  summarise(num_objects = sum(num_objects)) %>% 
  ungroup() %>% 
  ggplot(aes(x = year, y =num_objects))+
  geom_line()+
  scale_y_log10()
```

Here it's more obvious there were broadly two 'regimens' globally of objects released into space. 


Which countries released most objects into space over the whole period covered by the dataset?

```{r}
top_10 <- outer_space_objects %>% 
  filter(entity != "World", entity != "European Space Agency") %>% 
  count(entity, wt = num_objects, name = "amount") %>% 
  slice_max(order_by = amount, n = 10) %>% 
  pull(entity)
top_10
```

For these top 10 countries, we decided to look at trends over time:


```{r}
outer_space_objects %>% 
  filter(entity %in% top_10) %>% 
  ggplot(aes(x = year, y =num_objects, colour = entity))+
  geom_line()+
  scale_y_log10()+
  facet_wrap(~entity, scales = "free_y")
```

Tricia suggested we ordered these facets by the total number of objects released over the whole period. For this we used the `forcats` package, with a bit of trial and error. 

```{r}
# This use of fct_reorder looks like it's worked, but it hasn't
outer_space_objects %>%
  filter(entity %in% top_10) %>%
  mutate(entity = fct_reorder(entity, num_objects, .desc = TRUE)) %>%
  ggplot(aes(x = year, y =num_objects, colour = entity))+
  geom_line()+
  scale_y_log10()+
  facet_wrap(~entity, scales = "free_y")
```
This version looks initially that it's worked, but it hasn't. We can confirm this with the following: 


```{r}
# This confirms the ordering is wrong in the above
outer_space_objects %>%
  filter(entity %in% top_10) %>%
  group_by(entity) %>%
  summarise(num_objects = sum(num_objects)) %>%
  ungroup() %>%
  arrange(desc(num_objects))
```


So we use `group_by` but with `mutate`, not `summarise`, to create the `total_num_objects` column, which is then used by `fct_reorder` to create something that can be faceted correctly

```{r}
outer_space_objects %>%
  filter(entity %in% top_10) %>%
  group_by(entity) %>%
  mutate(total_num_objects = sum(num_objects)) %>%
  ungroup() %>%
  mutate(entity = fct_reorder(entity, total_num_objects, .desc = TRUE)) %>%
  ggplot(aes(x = year, y =num_objects, colour = entity))+
  geom_line()+
  scale_y_log10()+
  facet_wrap(~entity, scales = "free_y") + 
  labs(
    title = "Number of objects in outer space by top 10 countries",
    subtitle = "Arranged by total number of objects put in outer space"
  )


```

We can see, for example, that China, Japan and India's number of objects sent into space has been growing exponentially for a long time. The USA, and UK, appear to be having more of a recent exponential growth spurt. Russia's object release rate declined with the collapse of the USSR. 
