---
title: "Tidy Tuesday on Dr Who"
author: 
  - "Jon Minton"
  - "Nick Christofides"
date: "2023-11-29"
categories: [R, tidy tuesday, Dr Who]
---

## Tidy Tuesday challenge

First we load the packages 

The tidyverse equivalent of pacman is now pak.

The latest dataset is [here](https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-11-28), and the specific files to work. 



```{r}
library(tidyverse)

dta_list <- tidytuesdayR::tt_load(x = "2023-11-28")

dta_eps <- dta_list[["drwho_episodes"]]
dta_wrt <- dta_list[["drwho_writers"]]

```

Let's see how the viewship changed over time 

```{r}
dta_eps_season <- 
  dta_eps |> 
  group_by(season_number) |> 
  mutate(
    mean_viewers = mean(uk_viewers),
    mean_date = mean(first_aired)
    ) |> 
  ungroup()

dta_eps_season |> 
  ggplot(aes(x = first_aired, y = uk_viewers)) + 
  geom_point(colour = "grey") +
  geom_point(aes(x = mean_date, y = mean_viewers), size = 2.5) + 
  scale_x_date(breaks = "2 years", labels = \(x) format(x, "%Y")) +
  labs(
    x = "First aired",
    y = "UK Viewers (millions)",
    title = "Viewers over time for Dr Who",
    subtitle = "People don't watch TV like they used to..."
  ) +
  annotate("text", x = lubridate::make_date(2015), y = 10, label = "What happened here?!") +
  annotate("text", x = lubridate::make_date(2014), y= 8, label = "Smartphone strangling the TV from now", hjust = 0) + 
  stat_smooth(colour = "blue", se = FALSE) 
```

Let's now look at writers by season

```{r}
dta_eps_wrt <- 
  dta_eps |> 
    left_join(dta_wrt, by = "story_number") 

```

How many episodes by writer?

```{r}
dta_eps_wrt |> 
  group_by(writer) |>  
  summarise(
    n_written = n()
  ) |> 
  ungroup() |> 
  arrange(desc(n_written))
```
So Moffat wrote most episodes, then Davies, then Chibnall

And what about popularity by writer?

```{r}
dta_eps_wrt |> 
  group_by(writer) |>  
  mutate(
    n_written = n()
  ) |> 
  ungroup() |> 
  filter(n_written >= 5) |> 
  ggplot(aes(x = fct_reorder(writer, rating), y= rating)) + 
  geom_boxplot() + 
  coord_flip() + 
  labs(
    x = "Distribution of ratings",
    y = "Writer", 
    title = "Rating distribution by writer",
    subtitle = "Writers who wrote at least five episodes"
  )
```

When were the different writers active?

```{r}
major_writers_active <- 
  dta_eps_wrt |> 
    group_by(writer) |>  
    mutate(
      n_written = n()
    ) |> 
    ungroup() |> 
    filter(n_written >= 5) |> 
    group_by(writer) |> 
    summarise(
      started_writing = min(first_aired),
      finished_writing = max(first_aired),
      n_written = n_written[1]
    ) |> 
    ungroup() |> 
    mutate(
      yr_start = year(started_writing),
      yr_end = year(finished_writing)
    )

major_writers_active |> 
  arrange(started_writing)
```


Here we see the tenure of different major writers. Russell T Davies and Steven Moffatt are the major players.



## Coda

Neither of us know much about Dr Who! 

But hopefully we now know a bit more!




