---
title: "Tidy Tuesday: English Monarchs"
date: "2024-08-21"
categories: [tidy tuesday, monarchs, ggplot2, tidyverse, history, regex]
author:
    - Jon
    - Kate
    - Andrew
    - Brendan
    - Nic
    - Kennedy
    - Miryiam
    - Gatz
code-fold: true
---

The most recent Tidy Tuesday dataset was about [English kings](https://github.com/rfordatascience/tidytuesday/tree/master/data/2024/2024-08-20) (including female kings - there should be a word for that) and their consorts. 

For a change, Jon led the session. 

Load the data:

```{r}

library(tidyverse)


english_monarchs_marriages_df <- readr::read_csv(
  'https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-08-20/english_monarchs_marriages_df.csv'
  )
```

Data tidying - a lot of missing age information for older records, especially for the consorts:

```{r}
mar_tidy <-
  english_monarchs_marriages_df |> 
    mutate(
      king_age = str_remove_all(king_age, "\\?|\\–|\\(|\\)") %>% as.numeric(),
      consort_age = str_remove_all(consort_age, "\\?|\\–|\\(|\\)") %>% as.numeric(),
      year_of_marriage= str_remove_all(year_of_marriage, "\\?|\\–|\\(|\\)") %>% as.numeric()
    )
```

First plot: age of 'king' against age of consort:
```{r}
mar_tidy |> 
  ggplot(aes(x = king_age, y = consort_age)) +
  geom_point() +
  labs(title = "King Age vs Consort Age",
       x = "King Age",
       y = "Consort Age",
       subtitle = "Most kings were older than their consorts. QE2 Highlighted"
  ) +
  coord_equal(xlim = c(0, 65), ylim = c(0, 65)) +
  geom_abline(intercept =0, slope = 1, linetype = "dashed") +
  geom_vline(xintercept = 16, linetype = "dashed") + 
  geom_point(aes(x = king_age, y = consort_age), shape = 17, size = 3, colour = "purple", data = 
               mar_tidy |> filter(king_name == "Elizabeth II")
             )
```

We added a special royal purple triangle for QE2.

We wanted to know

> Were the 'kings' whose consorts were older than them actually Queens?


```{r}
mar_tidy |> 
  filter(king_age < consort_age) |> 
  select(king_name, consort_name, king_age, consort_age, year_of_marriage) |> 
  distinct() |> 
  arrange(king_age, consort_age)
```

Female 'kings' are over-represented but not the majority

Now difference between king and consort age by year of marriage 

```{r}
mar_tidy |> 
  mutate(age_diff = king_age - consort_age) |>
  ggplot(aes(x = year_of_marriage, y = age_diff)) +
  geom_point() +
  labs(title = "Difference between King and Consort Age by Year of Marriage",
       x = "Year of Marriage",
       y = "Difference between King and Consort Age"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth()
```

The age gap between kings and consorts appears to have been falling from around 1500. 
(Note: neither of King Charles' consorts are in the dataset as he was not a king at the time of either marriage)

Was there any obvious tendency for English Kings to have multiple wives, and - like Leo DeCaprio - keep partnering with young women as they age? 

```{r}
mar_tidy |> 
  group_by(king_name) |> 
  mutate(king_count = n()) |> 
  filter(king_count > 1) |> 
  filter(year_of_marriage > 1000) 
```

For many marriages the ages of the consorts weren't known (the records are from hundreds of years ago, so maybe that's to be expected). 

```{r}
mar_tidy |> 
  group_by(king_name) |> 
  mutate(king_count = n()) |> 
  filter(king_count > 1) |> 
  filter(!is.na(consort_age)) |>
  ggplot(aes(x = king_age, y = consort_age)) +
  geom_line() +
  geom_point() +
  coord_equal() + 
  coord_equal(xlim = c(0, 65), ylim = c(0, 65)) +
  geom_abline(intercept =0, slope = 1, linetype = "dashed") +
  facet_wrap(~king_name) + 
  labs(
    x = "Age of king",
    y = "Age of consort",
    title = "Ages of kings and consorts for multiple-marrying kings"
  )
```

Henry VIII is clearly an outlier in terms of the number of marriages he had. But there's no clear tendency for English kings to marry with Hollywood tendencies. 

Even where some kings have been married multiple times, the age of the consorts might only have been recorded one or zero times. 

For many of the older records honourifics (and in some cases dishonourifics) were applied to the name. These were identified by the king name having the word 'the' in them:

```{r}
mar_tidy |>
  filter(str_detect(king_name, " the ")) |> 
  pull(king_name) |>
  unique()
```

Better to be 'unready' or a 'martyr'? (probably best to be unready to be a martyr!)
